---
title: "proposal"
output: pdf_document
author: Ariel Estrin
date: "2026-03-16"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(quantmod)
library(readr)
library(rstan)
library(cmdstanr)
library(coda)
library(ggplot2)
library(bayesplot)

source("../func/simple_utils.R")

options(mc.cores = parallel::detectCores())
```

Oil markets are infamous for their volatility especially in times of increased uncertainty and geopolotical turmoil. This project aims to shed light on volatility o served in the global oil benchmark West Texas Intermediate (WTI) crude oil and/or Brent (COIL) under a uniquely Bayesian framework. After converting prices to log returns, I hope to estimate two models: a simple normal return model and a heavy-tailed Student model. Posterior inference will be computed using two different approaches: a basic Metropolis–Hastings MCMC sampler and perhaps an Hamiltonian Monte Carlo. I will compare these approaches using effective sample size, convergence diagnostics, and computational efficiency. The goal is to evaluate whether heavy-tailed models provide a better description of oil price fluctuations and to assess the efficiency gains of certain Bayesian computation methods. Daily WTI price data is obtained from the Federal Reserve Bank of St. Louis FRED database and Brent data from an open source csv file: `https://github.com/datasets/oil-prices/blob/main/data/brent-daily.csv`

## Scientific Question(s)

This project will address (some of) the following questions: 

- What is the estimated volatility of daily oil price returns?
- Do oil returns exhibit heavy tails relative to a normal distribution? 
- Does a more flexible model improve the fit to the observed return distribution? 
- How do different Bayesian computation methods compare in terms of efficiency and convergence?

## Bayesian Return Models

Two models will be considered for modeling oil price returns.

### Model 1: Normal Return Model (Baseline)

This model assumes that returns follow a normal distribution with constant volatility:

\[
r_t \sim \mathcal{N}(\mu, \sigma)
\]

**Priors:**

\[
\mu \sim \mathcal{N}(0, 10)
\]

\[
\sigma \sim \text{Half-Cauchy}(0, 2.5)
\]

### Mopdel 2 (heavy-tailed)

\
**Likelihood:**

\[
r_t \sim t_{\nu}(\mu, \sigma)
\]

where:  

- \(\mu\) is the mean return  
- \(\sigma\) is the scale parameter  
- \(\nu\) controls the heaviness of the tails  

**Priors:**

\[
\mu \sim \mathcal{N}(0, 10), \quad
\sigma \sim \text{Half-Cauchy}(0, 2.5), \quad
\nu \sim \text{Exponential}(1)
\]

## Data

```{r}
wti <- read.csv("../data/DCOILWTICO.csv") %>%
  rename(date = observation_date, price = DCOILWTICO) %>%
  mutate(date = as.Date(date),
         price = as.numeric(price),
         return = c(NA, diff(log(price)))) %>%
  na.omit()

brent <- read.csv("../data/brent-daily.csv") %>%
  mutate(Date = as.Date(Date),
         Price = as.numeric(Price),
         Return = c(NA, diff(log(Price)))) %>%
  na.omit()

head(wti)
head(brent)
```
